---
title: "ORF-level P-site quantification"
author: "Damon Hofman"
output: html_notebook
---

# Introduction

In this notebook, I calculate the number of translated ORFs based on the calculation of P-site per million values (similar to TPM values). I defined translated ORFs as ORFs that have a P-sites per million (PPM) value of > 1 in at least 5 samples. PPM is calculated in the same way as TPM, using raw P site counts as input instead of raw read counts. 

Input files:

* Sample metadata (Table S1A)
* Riboseq pipeline P site quantification output tables for canonical and non-canonical ORFs
* BED files containing all possible predicted P sites for canonical and non-canonical ORFs
* ORF reference table

Output files:
* cleaned up P site counts matrix
* table with translated ORFs
* figures relating to ORF translation (Fig 1I heatmap & Fig 1J bar plot)


```{r setup, include = F}
suppressPackageStartupMessages({
  library(tidyverse)
  library(rtracklayer)
  library(RColorBrewer)
  library(ComplexHeatmap)
  library(DESeq2)
  library(circlize)
  library(ggrepel)
  library(EnhancedVolcano)
  library(ggbeeswarm)
})
```

```{r}
colors <- c("#0073C2FF", "#EFC000FF", "#868686FF", "#DC0000FF", "#7AA6DCFF", "#1B1B1BFF", "#FF7F0EFF", "#17BCEFFF", "#009E73", "#CC79A7")
```

```{r functions, include = FALSE}
position_stack_and_nudge <- function(x = 0, y = 0, vjust = 1, reverse = FALSE) {
  ggproto(NULL, PositionStackAndNudge,
    x = x,
    y = y,
    vjust = vjust,
    reverse = reverse
  )
}

#' @rdname ggplot2-ggproto
#' @format NULL
#' @usage NULL
#' @noRd
PositionStackAndNudge <- ggproto("PositionStackAndNudge", PositionStack,
  x = 0,
  y = 0,

  setup_params = function(self, data) {
    c(
        list(x = self$x, y = self$y),
        ggproto_parent(PositionStack, self)$setup_params(data)
    )
  },

  compute_layer = function(self, data, params, panel) {
    # operate on the stacked positions (updated in August 2020)
    data = ggproto_parent(PositionStack, self)$compute_layer(data, params, panel)

    x_orig <- data$x
    y_orig <- data$y
    # transform only the dimensions for which non-zero nudging is requested
    if (any(params$x != 0)) {
      if (any(params$y != 0)) {
        data <- transform_position(data, function(x) x + params$x, function(y) y + params$y)
      } else {
        data <- transform_position(data, function(x) x + params$x, NULL)
      }
    } else if (any(params$y != 0)) {
      data <- transform_position(data, function(x) x, function(y) y + params$y)
    }
    data$nudge_x <- data$x
    data$nudge_y <- data$y
    data$x <- x_orig
    data$y <- y_orig

    data
  },

  compute_panel = function(self, data, params, scales) {
      ggproto_parent(PositionStack, self)$compute_panel(data, params, scales)
  }
)
```


# Pre-processing the P site counts table

```{r}
# Input file locations

# Canonical P-site counts matrix
MANE_mat_location <- "/hpc/pmc_vanheesch/projects/Damon/Medulloblastoma_analyses_final/riboseq/data/processed/psite_quantification/Medullo_all_Psites_quant_MANE/Medullo_all_Psites_quant_MANE_new_allsamples_psites.txt"
# Non-canonical P-site counts matrix
gencode_mat_location <- "/hpc/pmc_vanheesch/projects/Damon/Medulloblastoma_analyses_final/riboseq/data/processed/psite_quantification/Medullo_all_Psites_quant_GENCODEp1_plus_ORFEOME/Medullo_all_Psites_quant_GENCODEp1_plus_ORFEOME_new_allsamples_psites.txt"

# Metadata file
metadata_location <- "/hpc/pmc_vanheesch/projects/Damon/Medulloblastoma_analyses_final/01.26.2023_Table_S1.xlsx"  # Change to generic table S1 location

# ORF reference table 
ref_table_location <- "/hpc/pmc_vanheesch/projects/Damon/Medulloblastoma_analyses_final/annotation_files/all_ORFs_ref_table.csv"  # Table containing gene_id, tx_id, orf_id, gene_name, biotype & 'canonical' columns, used to match genes to transcripts and ORFs
noncanonical_orf_data_location <- "/hpc/pmc_vanheesch/projects/Damon/Medulloblastoma_analyses_final/annotation_files/noncanonical_ORFs_data.csv"   # = Table S1L
```

```{r}
# Load metadata table S1A
sample_metadata <- readxl::read_xlsx(metadata_location, 
  sheet = 1, 
  col_names = T, 
  skip = 2, 
  .name_repair = "universal")

# Filter samples with both RNAseq and riboseq data
sample_metadata_complete <- sample_metadata[which(
  !is.na(sample_metadata$RNASeq_file_names) &
    !is.na(sample_metadata$RiboSeq_file_names..final.)), ]
```

```{r}
# Load ORF reference table
ref_table <- read.csv(ref_table_location)
noncanonical_orf_data <- read.csv(noncanonical_orf_data_location)
```

```{r}
# Load P sites data
MANE_mat <- read.table(MANE_mat_location, check.names = F)
gencode_mat <- read.table(gencode_mat_location, check.names = F)

# Combine P MANE and gencode P sites data
counts_psites <- data.frame(rbind(MANE_mat, gencode_mat), check.names = F)
# colnames(counts_psites)[which(colnames(counts_psites) == "S15_1824_S1_11_R1")] <- "S15_1824_JP10031_S11"

counts_psites <- counts_psites[, grep("HHT", colnames(counts_psites), invert = T)]  # Remove HHT samples

# Fix rownames
rownames(counts_psites) <- gsub("\\..{1,2}$","", rownames(counts_psites))  # Remove transcript version for MANE transcripts

# Give canonical ORFs (transcripts) "_CDS" suffix
# canonical_ind <- which(rownames(counts_psites) %in% subset(ref_table, ref_table$canonical == "Canonical")$tx_id)

canonical_ind <- which(!(rownames(counts_psites) %in% noncanonical_orf_data$orf_id))

rownames(counts_psites)[canonical_ind] <- paste0(rownames(counts_psites)[canonical_ind], "_CDS")
```

```{r}
# Match P-site counts columns to metadata rows

# Extract the column names from rna_counts that match the RNASeq_file_names in sample_metadata_celllines
match_counts <- unique(str_extract(sample_metadata_complete$RiboSeq_file_names..final., str_c(colnames(counts_psites), collapse = "|")))

# Subset rna_counts based on the matched columns
counts_psites <- counts_psites[, match_counts]

# Find matching metadata
metadata_matches <- grep(paste(match_counts, collapse = "|"), sample_metadata_complete$RiboSeq_file_names..final.)
sample_metadata_complete <- sample_metadata_complete[metadata_matches, ]

colnames(counts_psites) <- sample_metadata_complete$Sample_ID
```


```{r}
write.csv(counts_psites, "/hpc/pmc_vanheesch/projects/Damon/Medulloblastoma_analyses_final/riboseq/results/tables/S1_riboseq_P_sites_counts.csv", quote = F)
```


# Identification of translated ORFs

```{r load ORF lengths}
# Load predicted P sites for each ORF or transcript to calculate ORF lengths
bed_canonical_p_all <- data.table::fread(
  "/hpc/pmc_vanheesch/projects/Damon/Medulloblastoma_analyses_final/riboseq/data/processed/psite_quantification/Medullo_all_Psites_quant_MANE/bedfiles/MANE.GRCh38.v1.0.ensembl_genomic.nochr.gtf_psites_p0.sorted.bed",
  col.names = c("chrom", "start", "end", "ref_id", "frame", "strand")) %>%
  subset(!grepl("pATG|pST", .$frame)) 

bed_novel_p_all <- data.table::fread(
  "/hpc/pmc_vanheesch/projects/Damon/Medulloblastoma_analyses_final/riboseq/data/processed/psite_quantification/Medullo_all_Psites_quant_GENCODEp1_plus_ORFEOME/bedfiles/GENCODEpI_plus_ORFEOME3.bed.orfs_new.gtf_psites_p0.sorted.bed",
  col.names = c("chrom", "start", "end", "ref_id", "frame", "strand")) %>%
  subset(!grepl("pATG|pST", .$frame)) 

ORF_lengths <- rbind(bed_canonical_p_all, bed_novel_p_all) %>%
  # rownames_to_column("ref_id") %>%
  group_by(ref_id) %>%
  summarize(
    n_codons = n()/3, 
    length = n(), 
    length_kb = n()/1000)

ORF_lengths$ref_id <- gsub("\\..{1,2}$","", ORF_lengths$ref_id)

canonical_ind <- which(!(ORF_lengths$ref_id %in% noncanonical_orf_data$orf_id))
ORF_lengths$ref_id[canonical_ind] <- paste0(ORF_lengths$ref_id[canonical_ind], "_CDS")
ORF_lengths <- ORF_lengths %>%
  column_to_rownames("ref_id")
```


## Selection of translation cutoff

In the following section, I calculate the P-sites per million (ppm) of each ORF, which is calculated the same way as transcripts per million (TPM) values, but uses P-sites counts instead of transcript counts.
6567 ORFs have a mean ppm < 1. 

```{r}
row_indices <- match(rownames(ORF_lengths), rownames(counts_psites))
# any(is.na(row_indices)) # FALSE; check if any rownames of ORF_lengths failed to map to rownames of counts_psites
ORF_lengths_match <- ORF_lengths[order(row_indices),]

all(rownames(ORF_lengths_match) == rownames(counts_psites))

lengths_kb <- ORF_lengths_match$length_kb

rpk <- counts_psites / lengths_kb
pm_scalingfactors <- colSums(rpk)/1000000
ppm <- t(t(rpk) / pm_scalingfactors)
ppm_log2 <- log2(ppm + 0.01)

table(rowMeans(ppm) < 1)
low_orfs <- rownames(ppm)[which(rowMeans(ppm) < 1)]
```

To select a cutoff, I plot the density of log2-transformed ppm values in each sample, and choose a cutoff between the low peak and the higher peak, which is around log2(ppm) = 0, or a ppm of 1. 

```{r}
dens = density(ppm_log2)
plot(dens)
# threshold = identify(dens)
```


```{r}
cutoff = 0
threshold <- which(abs(dens$x-cutoff)==min(abs(dens$x-cutoff)))  # Find dens$x observation closest to
# threshold=120
plot(dens)
abline(v = dens$x[threshold])
```


## Calculating the total number translated ORFs

Next, I make a list of ORFs to keep based on a ppm > 1 in at least 5 samples. 
Based on these criteria, 5582 ORFs are removed.

```{r select ORFs from low cluster to keep}
# Make logical table indicating whether ORF is above or below cutoff
mat_cutoff <- ifelse(ppm > 1, 1, 0)

mat_cutoff_summary <- data.frame(number_samples = rowSums(mat_cutoff)) %>%
  mutate(canonical = ifelse(rownames(.) %in% noncanonical_orf_data$orf_id, "non-canonical", "canonical"))
  
mat_cutoff_extended <- mat_cutoff_summary %>%
  rownames_to_column("orf_id") %>%
  # left_join(ref_table, by = "orf_id")  %>%
  left_join(noncanonical_orf_data, by = "orf_id")

ORFs_keep <- rownames(subset(mat_cutoff_summary, number_samples >= 5))

# write.csv(mat_cutoff_summary, "/hpc/pmc_vanheesch/projects/Damon/Medulloblastoma_analyses/02_results/notebooks/orf_quantification/ORF_translation_summary.csv")

write.csv(mat_cutoff_extended, "/hpc/pmc_vanheesch/projects/Damon/Medulloblastoma_analyses_final/riboseq/results/tables/S1_ORF_translation_extended.csv", row.names = F)
```


## Calculating the number of translated canonical CDSs

```{r}
# Select canonical CDSs from logical cutoff table
mat_cutoff_canonical <- mat_cutoff[which(rownames(mat_cutoff) %in% rownames(subset(mat_cutoff_summary, canonical == "canonical"))), ]

# Calculate number and percentage of translated CDSs per samples
# orfs_per_sample_canonical <- data.frame(Sample_ID = colnames(mat_cutoff_canonical),
#                                         num_orfs = colSums(mat_cutoff_canonical)) %>%
#   mutate(perc_canonical = num_orfs / nrow(subset(ref_table, canonical == "Canonical"))) %>%
#   left_join(sample_metadata_complete)

orfs_per_sample_canonical <- data.frame(Sample_ID = colnames(mat_cutoff_canonical),
                                        num_orfs = colSums(mat_cutoff_canonical)) %>%
  left_join(sample_metadata_complete)

max(orfs_per_sample_canonical$num_orfs)
min(orfs_per_sample_canonical$num_orfs)
mean(orfs_per_sample_canonical$num_orfs)
```


```{r}
# Plot with number of translated canonical CDSs per sample
orfs_per_sample_canonical[grep(pattern = "Tissue", x = orfs_per_sample_canonical$Sample_Type), ]$Sample_Type <- "Tissue"

ggplot(data = orfs_per_sample_canonical, aes(y = num_orfs, x = Sample_ID)) +
  geom_bar(stat = "identity", fill = colors[1]) +
  geom_hline(yintercept = mean(orfs_per_sample_canonical$num_orfs), linetype = 2) +
  geom_text(data = data.frame(Type = "Cell line"), aes(label = round(mean(orfs_per_sample_canonical$num_orfs))), x = -1, y = mean(orfs_per_sample_canonical$num_orfs), 
            inherit.aes = FALSE, color = "grey20", hjust = .6, vjust = .5, size = 8 / .pt) +
  coord_cartesian(clip = "off", ylim = c(1, NA)) +
  scale_y_continuous(expand = c(0,0)) +
  labs(y = "Translated canonical proteins (ppm > 1)", x = "Samples") +
  facet_grid(~ Sample_Type, scales = "free_x", space = "free_x") +
  
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 3))


ggsave("/hpc/pmc_vanheesch/projects/Damon/Medulloblastoma_analyses_final/riboseq/results/plots/1C_num_translated_CDS.pdf", 
       width = unit(4, "cm"),
       height = unit(4, "cm"))

```


# Normalized P site counts in all samples

Figures 1I and 1J

First, I normalize all the ORF P site counts together using DESeq2. 

```{r}
# Get coldata
coldata_select <- sample_metadata_complete[match(colnames(counts_psites), sample_metadata_complete$Sample_ID), ]
rownames(coldata_select) = coldata_select$Sample_ID

coldata_select[which(coldata_select$Sample_Type == "Tissue_autopsy"), "Sample_Type"] <- "Tissue" 

coldata_select[is.na(coldata_select$MYC.group), "MYC.group"] <- "Unknown"
coldata_select$MYC.group <- factor(coldata_select$MYC.group)

# Fix metadata
colnames(coldata_select) <- gsub(" ", "_", colnames(coldata_select))
colnames(coldata_select) <- gsub("-", "_", colnames(coldata_select))
```

```{r}
dds <- DESeq2::DESeqDataSetFromMatrix(countData = counts_psites, colData = coldata_select, design = ~ Sample_Type + MYC.group)
dds <- DESeq(dds)
vsd <- DESeq2::vst(dds, blind = F)

mat_dds <- data.frame(counts(dds, normalized = T))
mat_vsd <- assay(vsd)
```

Next, I select only the normalized counts for noncanonical ORF P sites.
```{r}
dds_NC <- dds[ORFs_keep, ]
dds_NC <- dds_NC[!grepl("_CDS", rownames(dds_NC)),]
vsd_NC <- vsd[rownames(dds_NC), ]
# mat_vsd_NC <- assay(vsd_NC)
mat_vsd_NC <- assay(vsd)[rownames(dds_NC), ]

```


## Heatmap of translated noncanonical ORFs

```{r}
# Get MYC expression levels
ind_MYC <- ref_table[grep("ENSG00000136997", ref_table$gene_id),]$orf_id 
MYC_vsd <- as.numeric(mat_vsd[ind_MYC, ])
MYC_dds <- as.numeric(mat_dds[ind_MYC, ])

ind_MYCN <- ref_table[grep("ENSG00000134323", ref_table$gene_id),]$orf_id 
MYCN_vsd <- as.numeric(mat_vsd[ind_MYCN, ])
MYCN_dds <- as.numeric(mat_dds[ind_MYCN, ])

```


```{r}
# Set heatmap colors
myCol <- colorRampPalette(rev(brewer.pal(n = 7, name = "RdYlBu")))(100)
myBreaks <- seq(0, 8, length.out = 100)
colors <- c("#0073C2FF", "#EFC000FF", "#868686FF", "#DC0000FF", "#7AA6DCFF", "#1B1B1BFF", "#FF7F0EFF", "#17BCEFFF", "#009E73", "#CC79A7")
```


```{r}
# Column annotation

# Make column annotation df
col_ann <- data.frame(
  Sample_Type = dds_NC$Sample_Type, 
  # MYC_status = dds_NC$MYC_hi_lo,
  MYC_level = MYC_vsd,
  # MYCN_level = MYCN_vsd,
  row.names = colnames(assay(vsd_NC)))

col_ann$Sample_Type <- factor(col_ann$Sample_Type, levels = sort(unique(col_ann$Sample_Type)))
# col_ann$MYC_status <- factor(col_ann$MYC_status, levels = sort(unique(col_ann$MYC_status)))

# Set annotation colors
anno_cols <- list(
  Sample_Type = setNames(c(brewer.pal(3, "Pastel1")[1:length(unique(col_ann$Sample_Type))]), 
                     c("Cell line", "Tissue")), 
  # MYC_status = setNames(c(brewer.pal(length(unique(col_ann$MYC_status)), "Set1")[1:2]),
                                  # unique(col_ann$MYC_status)),
  MYC_level = colorRamp2(
    breaks = seq(min(MYC_vsd), max(MYC_vsd), length.out = 100),
    myCol)
  # MYCN_level = colorRamp2(
  #   breaks = seq(min(MYCN_vsd), max(MYCN_vsd), length.out = 100),
  #   myCol)
  )

# Make column HeatmapAnnotation
ha <- HeatmapAnnotation(
  df = col_ann, 
  which = "column", 
  col = anno_cols, 
  show_annotation_name = F, 
  simple_anno_size = unit(0.2, "cm"), 
    annotation_legend_param = list(direction = "vertical", ncol = 2, labels_gp = gpar(fontsize = 6), title_gp = gpar(fontsize = 8, fontface = "bold")))
```


```{r}
# Row annotations

ref_table_NC <- ref_table[!grepl("_CDS", rownames(ref_table)),]

# Make row annotation df
df_row = data.frame(
  orf_id = rownames(mat_vsd_NC)) %>%
  left_join(ref_table_NC[,c("orf_id", "biotype")]) %>%
  column_to_rownames("orf_id")

ref_table_NC_summary <- ref_table_NC %>%
  group_by(biotype) %>%
  summarize(count = n())

df_row_summary <- df_row %>%
  group_by(biotype) %>%
  summarize(count = n()) %>%
  mutate(biotype_count = paste0(biotype, " (n = ", count, ")"))

df_row_2 <- rownames_to_column(df_row, "orf_id") %>%
  left_join(df_row_summary) %>%
  mutate(biotype = paste0(biotype, " (n = ", count, ")")) %>%
  column_to_rownames("orf_id") %>%
  select(biotype)

df_row_2$biotype <- factor(
  df_row_2$biotype,
  levels = c(df_row_summary[order(df_row_summary$count, decreasing = T), ]$biotype_count))



# Set row annotation colors
col_row <- list(
  biotype = setNames(colors[1:length(unique(df_row_2$biotype))], 
                     unique(df_row_2$biotype)))
# Make row HeatmapAnnotation
row_ha = HeatmapAnnotation(
  which = "row",
  df = df_row_2,
  annotation_name_gp = gpar(fontsize = 6), 
  show_annotation_name = F,
  col = col_row, 
  show_legend = F)
```


```{r}
# Heatmap params
hm <- Heatmap(
  
  mat_vsd_NC, 
  
  name = "DESeq2 norm. transformed counts", 
  
  col = colorRamp2(myBreaks, myCol),

  # Parameters for expression color-bar legend
  heatmap_legend_param = list(
    color_bar = "continuous",
    title_position = "topcenter", 
    legend_width = unit(4, "cm"), 
    title_gp = gpar(fontsize = 8, fontface = "bold"), 
    labels_gp = gpar(fontsize = 8), 
    direction = "horizontal"),
  
  # Row parameters
  cluster_rows = T, 
  show_row_names = F, 
  row_names_gp = gpar(fontsize = 6),
  row_title_gp = gpar(fontsize = 6, fontface = "bold"),
  row_title_rot = T,
  show_row_dend = F,
  cluster_row_slices = F,
  row_split = df_row_2$biotype, 

  
  # Column parameters
  column_split = factor(col_ann$Sample_Type, levels = unique(col_ann$Sample_Type)), 
  show_column_names = F, 
  show_column_dend = T, 
  column_title_gp = gpar(fontsize = 8, fontface = "bold"),

  # Set column order
  cluster_columns = T, 
  cluster_column_slices = T,

  # Column and row annotations
  top_annotation = ha,
  left_annotation = row_ha,
  use_raster = T
  # raster_resize_mat = TRUE
  )

pdf("/hpc/pmc_vanheesch/projects/Damon/Medulloblastoma_analyses_final/riboseq/results/plots/1I_MYClvls_new.pdf", width = 4, height = 4)
ht <- draw(
  hm,
  heatmap_legend_side = "bottom",
  annotation_legend_side = "right",
  column_title_gp = gpar(fontsize = 16))
dev.off()
```


## Average P sites per biotype

```{r}
counts_summary <- data.frame(mean = rowMeans(counts(dds_NC, normalized = T))) %>%
  rownames_to_column("orf_id") %>% 
  left_join(ref_table[,c("orf_id", "biotype")]) %>%
  column_to_rownames("orf_id") %>%
  mutate(biotype = ifelse(is.na(.$biotype), "Canonical", .$biotype))

ORF_df <- data.frame(orf_id = ORFs_keep) %>%
  left_join(ref_table) %>%
  mutate(biotype = ifelse(is.na(.$biotype), "Canonical", .$biotype))

ORF_df_summary <- ORF_df %>%
  group_by(biotype) %>%
  summarize(count = n())

ORF_df_summary[is.na(ORF_df_summary$biotype), "biotype"] <- "Canonical"

counts_summary$biotype <- factor(counts_summary$biotype, levels = c(ORF_df_summary[order(ORF_df_summary$count, decreasing = T), ]$biotype))
counts_summary$biotype <- droplevels(counts_summary$biotype)

ggplot(data = counts_summary, aes(x = biotype, y = mean, fill = biotype, color = biotype)) + 
  geom_quasirandom(size = 0.1, color = "grey20") +
  # coord_flip(ylim = quantile(counts_summary$mean, c(0, 0.95))) +

  scale_y_continuous(trans = "log2") +

  coord_flip() +
  geom_boxplot(alpha = 0.7, outlier.shape = NA, color = 1) +
    scale_x_discrete(limits = rev(levels(counts_summary$biotype))) +
  # coord_cartesian(ylim = quantile(counts_summary$mean, c(0, 0.95))) +

  # viridis::scale_color_viridis(discrete = T, option = "C", begin = 0.1, end = 1) +
  # viridis::scale_fill_viridis(discrete = T, option = "C", begin = 0.1, end = 1) +
  scale_fill_manual(values = setNames(colors[1:length(unique(df_row$biotype))], 
                     unique(df_row$biotype))) +

  # stat_summary(fun = "median", colour = "red", size = 0.5,
  #              geom = "hline", aes(yintercept = after_stat(y))) +
  labs(x = "ORF type", y = "Norm. P-site counts") +
  theme_classic() +
  theme(legend.position = "none", axis.text.x = element_text(size = 6))

ggsave(
  filename = "/hpc/pmc_vanheesch/projects/Damon/Medulloblastoma_analyses_final/riboseq/results/plots/1J_log2.pdf", device = "pdf", path = , width = unit(3, "cm"), height = unit(3, "cm"))
```



