---
title: "Identifying ORFs differentially translated between MYC high and MYC low samples (cell lines only)"
output: html_notebook
---

# Description
This R Notebook is used to analyze differentially translated open reading frames (ORFs) in MYC high and MYC low cell lines. The analysis includes loading metadata, filtering samples, running DESeq2, creating volcano plots, and generating PCA plots.

## Input files:
1. Metadata file: 01.26.2023_Table_S1.xlsx
2. ORF reference table: all_ORFs_ref_table.csv
3. Riboseq P-sites counts: S1_riboseq_P_sites_counts.csv (generated with 03_riboseq_p_sites_quantification.Rmd)

## Output files:
1. ORF level differential translation analysis results: ORF_level_diff_transl_analysis_results.csv

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

suppressPackageStartupMessages({
  library(tidyverse)
  library(RColorBrewer)
  library(DESeq2)
  library(ggrepel)
  library(EnhancedVolcano)
})
```

# Set file locations

```{r file-locations}
# Metadata file
metadata_location <- "Table_S1.xlsx"

# ORF reference table
ref_table_location <- "/annotation_files/all_ORFs_ref_table.csv"
```

# Load and filter metadata

```{r load-metadata}
# Load metadata table S1A
sample_metadata <- readxl::read_xlsx(metadata_location, 
  sheet = 1, 
  col_names = T, 
  skip = 2, 
  .name_repair = "universal")

# Filter samples with both RNAseq and riboseq data
sample_metadata_complete <- sample_metadata[which(
  !is.na(sample_metadata$RNASeq_file_names) &
    !is.na(sample_metadata$RiboSeq_file_names..final.)), ]

sample_metadata_filtered <- subset(sample_metadata_complete, Sample_Type == "Cell line")
```

# Load data and match P-site counts columns

```{r load-data}
counts_psites <- read.csv("/riboseq/results/tables/S1_riboseq_P_sites_counts.csv", row.names = 1, check.names = F)
ref_table <- read.csv(ref_table_location)
```

```{r match-columns}
# Extract the column names from rna_counts that match the RNASeq_file_names in sample_metadata_celllines
match_counts <- unique(str_extract(sample_metadata_filtered$Sample_ID, str_c(colnames(counts_psites), collapse = "|")))

# Reorder the columns of P-site counts data frame to match the order of metadata
counts_psites_ordered <- counts_psites[, match_counts]

# Find matching metadata
metadata_matches <- grep(paste(match_counts, collapse = "|"), sample_metadata_filtered$Sample_ID)
sample_metadata_matched <- sample_metadata_filtered[metadata_matches, ]

colnames(counts_psites) <- sample_metadata_filtered$Sample_ID

# Fix metadata
coldata_select <- sample_metadata_filtered
colnames(coldata_select) <- gsub(" ", "_", colnames(coldata_select))
colnames(coldata_select) <- gsub("-", "_", colnames(coldata_select))
```


# Differential translation analysis (Table S1K)

```{r, eval = FALSE}
# Run DESeq2 to analyze differential translation
dds <- DESeqDataSetFromMatrix(countData = counts_psites_ordered,
                              colData = sample_metadata_matched,
                              design = ~ MYC.group)

# Run DESeq2
dds <- DESeq(dds)

# Transform DESeq2 output
vsd <- DESeq2::vst(dds, blind = F)

# Store dds counts and vsd counts in matrices
mat_dds <- data.frame(counts(dds, normalized = T))
mat_vsd <- assay(vsd)

# Get DESeq2 results
res <- results(dds, contrast = c("MYC.group", "MYC high", "MYC low"), alpha = 0.01)

# Make readable data frame
res_df <- data.frame(res) %>%
  rownames_to_column("orf_id") %>%
  left_join(ref_table) %>%
  mutate(orf_name = paste0(.$gene_name, "_", .$biotype))

res_df[which(is.na(res_df$gene_name)), "orf_name"] <- res_df[which(is.na(res_df$gene_name)), "orf_id"]

# Save ORF level differential translation analysis results
write.csv(res_df, "ORF_level_diff_transl_analysis_results.csv")
```

# Visualization: Volcano Plot (Fig 1K)

```{r volcano-plot, eval = FALSE}
# Create a volcano plot for the differential translation analysis results of non-canonical ORFs

# Select non-canonical ORF results
res_NC_df <- subset(res_df, canonical == "Non-canonical")
res_2 <- res_NC_df

# Select top 5 downregulated ORFs (log2FC < -2, based on padj)
top5down <- res_NC_df %>%
  subset(log2FoldChange < 2) %>%
  arrange(.$padj) %>%
  head(5)

# Select top 5 upregulated ORFs (log2FC < 2, based on padj)
top5up <- res_NC_df %>%
  subset(log2FoldChange > 2) %>%
  arrange(.$padj) %>%
  head(5) 

# Combine top 5 up and down ORFs
top <- rbind(top5down, top5up)

# Remove trailing '_biotype' part from ORF names to label
top$orf_name <- gsub("_.*$", "", top$orf_name)

# Remove the top 10 differentially regulated ORFs from results dataset
res_2 <- res_2[-which(res_2$orf_id %in% top$orf_id), ]

# Set ORF name of the rest of the ORFs to NA (so they won't be labeled)
res_2$orf_name <- NA

# Add results of the top 10 ORFs, including their ORF names (so only these will be labeled)
res_2 <- rbind(res_2, top)

# Make custom keyvals table
keyvals <- data.frame(res_2) %>%
  select(c("orf_id", "biotype", "padj"))

cols_biotype <- brewer.pal(6, "Set1")

# Set color of points by ORF biotype
keyvals[which(keyvals$padj < 0.01 & keyvals$biotype == "lncRNA"), "color"] <- cols_biotype[1]
keyvals[which(keyvals$padj < 0.01 & keyvals$biotype == "uORF"), "color"] <- cols_biotype[2]
keyvals[which(keyvals$padj < 0.01 & keyvals$biotype == "uoORF"), "color"] <- cols_biotype[3]
keyvals[which(keyvals$padj < 0.01 & keyvals$biotype == "dORF"), "color"] <- cols_biotype[4]
keyvals[which(keyvals$padj < 0.01 & keyvals$biotype == "intORF"), "color"] <- cols_biotype[5]
keyvals[which(keyvals$padj < 0.01 & keyvals$biotype == "doORF"), "color"] <- cols_biotype[6]
keyvals$name <- keyvals$biotype
keyvals[which(is.na(keyvals$color)), "name"] <- NA
keyvals[which(is.na(keyvals$color)), "color"] <- "grey"

# Make named list from keyvals table (needed for volcanoplot)
keyvals_list <- keyvals$color
names(keyvals_list) <- keyvals$name

# Plot the volcanoplot
plt <- EnhancedVolcano(
  res_2,
  lab = res_2$orf_name, legendLabels = c("NS", "NS", "NS", "p < 0.01 & log2FC > 2 | < -2"),
  colCustom = keyvals_list,

  x = "log2FoldChange", 
  y = "padj", 
  title = NULL,
  subtitle = NULL, colAlpha = 1, col = c("grey", "grey", "grey", "red"), pCutoff = 0.01, FCcutoff = 0, xlim = c(-12, 12), labSize = 3, labFace = "bold", max.overlaps = 10, drawConnectors = T, pointSize = 1, lengthConnectors = unit(0, "mm"), axisLabSize = 15, legendLabSize = 10)
plt

```

# Visualization: PCA Plot

```{r pca-plot, eval = FALSE}
# Create a PCA plot for the P-site counts data
pcaData <- plotPCA(vsd, intgroup = c("MYC.group"), returnData = T) 
  
percentVar <- round(100 * attr(pcaData, "percentVar"))

ggplot(pcaData, aes(PC1, PC2, color=MYC.group)) +
  scale_color_brewer(palette = "Set1") +
  geom_point(size=3) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
                                                           panel.background = element_blank(), axis.line = element_line(colour = "black"))+ theme(aspect.ratio=1) +
  labs(color = "MYC level", title = "P-site counts (cell lines)")
```

# Visualization: ASNSD1-uORF normalized counts plot (Fig 4A)

```{r ASDURF-counts}
# Create a dotplot of normalized ASNSD1-uORF P site counts in MYC high and MYC low samples

ASDURF_counts <- plotCounts(dds = dds, gene = "ENST00000607829_CDS", intgroup = "MYC.group", returnData = T, normalized = T)

ASDURF_plt <- ggplot(data = ASDURF_counts, aes(y = count, x = MYC.group, color = MYC.group)) +
  scale_color_brewer(palette = "Set1") +
  labs(title = "ASDURF P-site counts", x = NULL, y = "P-site counts") +
  geom_beeswarm() +
  theme_classic() +
  ggpubr::stat_compare_means(comparisons = list(c("MYC high", "MYC low")), label = "p.format", ref.group = "MYC low", label.y = max(ASDURF_counts$count)* 1.1)
ASDURF_plt

# Save the plot
ggsave(plot = ASDURF_plt,
       filename = "S4A_ASDURF_Riboseq.pdf", device = "pdf", path = , width = unit(6, "cm"), height = unit(6, "cm"))

```

# End of the R Notebook