---
title: "Medullo Riboseq Periodicity Plots"
output: html_notebook
author: "Damon Hofman"
---

# Description
This R notebook performs a periodicity analysis on medulloblastoma RiboseQC output data and generates Figures S1D and S1E.

## Input files:

1. 01.26.2023_Table_S1.xlsx: metadata table

## Output files:

1. Medullo_merged_periodicity_new.pdf: periodicity plot for overall stats
2. framepref_allreads_allsamples.csv: stats on the frame preference for all samples

# Load required libraries

```{r setup, include = FALSE}
library(RiboseQC)
library(tidyverse)
library(gridExtra)
```

# Load and preprocess metadata

```{r}
# Load metadata table S1A
sample_metadata <- readxl::read_xlsx(
  "Table_S1.xlsx", 
  sheet = 1, 
  col_names = T, 
  skip = 2, 
  .name_repair = "universal")

# Filter samples with both RNAseq and riboseq data
sample_metadata_complete <- sample_metadata[which(
  !is.na(sample_metadata$RNASeq_file_names) &
    !is.na(sample_metadata$RiboSeq_file_names..final.)), ]
```

# Gather stats per sample

```{r}
ribo_files <- list.files("/riboseq/data/processed/RiboseQC/", "results_RiboseQC_all", recursive = T, full.names = T)
names(ribo_files) <- basename(dirname(ribo_files))

ribo_files <- ribo_files[grep("HHT", names(ribo_files), invert = T)]

# Only select riboseqc files that are in the metadata
matches <- unique(str_extract(sample_metadata_complete$RiboSeq_file_names..final., str_c(names(ribo_files), collapse = "|")))
ribo_files <- ribo_files[matches]
```


```{r}
framepref_allreads_allsamples <- data.frame()

for(i in 1:length(ribo_files)){
  ribo_file <- ribo_files[i]
  sample <- basename(dirname(ribo_file))
  
  print(paste0("Processing sample ", i, " of ", length(ribo_files)))
  
  # Load riboseqc data
  load(ribo_file)
  data <- res_all
  profile_type = "profiles_P_sites"
  res = "P_sites_subcodon"
  comp = "nucl"
  signal <- data[[profile_type]][[res]][[comp]]
  
  signal <- data$profiles_fivepr$five_prime_subcodon$nucl
  rl_ok <- names(signal)  # get available read lengths
  rl_ok <- c(rl_ok[1], as.character(sort(as.integer(rl_ok[-1], decreasing=FALSE))))  # sort read lengths in ascending order and "all" at the beginning
  
  data_bar = list()
  
  for (rl in rl_ok) {
    # get read length data and format
    x <- data.frame(signal[[rl]], check.names = F)
    x.m <- data.frame(value = colSums(x))
    x.m$pos <- seq_len(length(x))
    x.m$frame <- paste0("frame ", rep(c(3,1,2),67)[seq_along(x)])
    
    # add read length data to list
    data_bar[[rl]] <- x.m
  }
  
  
  # Get read length for max frame preference in sample
  summary_P_sites_sample <- data.frame(data$summary_P_sites)
  rl <- subset(summary_P_sites_sample, read_length %in% c("28", "29", "30")) %>%
  dplyr::slice(which.max(frame_preference)) %>%
  dplyr::select(read_length)
  rl <- as.character(rl)
  
  # rl <- ifelse("29" %in% rl_ok, "29", "all")
  # rl <- ifelse(rl %in% rl_ok, "29", "all")
  
  
  # Get periodicity data for max read length
  x.m <- data_bar[[rl]]
  
  res_breaks <- c(1, 26, 51, 84, 117, 150, 176, 200)
  res_labels <- c("TSS", "", "start\ncodon", "", "", "stop\ncodon", "", "TES")
  res_x_lab <- "\nposition (nucleotide resolution)"
  
  x.m$frame<-as.character(x.m$frame)
  x.m$frame[x.m$frame=="frame 2"]<-"Frame 0"
  x.m$frame[x.m$frame=="frame 1"]<-"Frame 2"
  x.m$frame[x.m$frame=="frame 3"]<-"Frame 1"
  x.m$frame<-gsub(x.m$frame,pattern = "F",replacement = "f")
  x.m$frame<-factor(x.m$frame,levels=c("frame 0","frame 1","frame 2"))
  
  framepref_allreads <- data_bar[[rl]] %>%
    group_by(frame) %>%
    summarize(tot_reads = sum(value)) %>%
    mutate(perc = tot_reads/sum(tot_reads))
    
  framepref_allreads$frame<-as.character(framepref_allreads$frame)
  framepref_allreads$frame[framepref_allreads$frame=="frame 2"]<-"Frame 0"
  framepref_allreads$frame[framepref_allreads$frame=="frame 1"]<-"Frame 2"
  framepref_allreads$frame[framepref_allreads$frame=="frame 3"]<-"Frame 1"
  framepref_allreads$frame<-gsub(framepref_allreads$frame,pattern = "F",replacement = "f")
  framepref_allreads$frame<-factor(framepref_allreads$frame,levels=c("frame 0","frame 1","frame 2"))
  framepref_allreads$sample<-sample
  framepref_allreads$rl <- rl
  framepref_allreads_allsamples <- rbind(framepref_allreads_allsamples, framepref_allreads)
  
  print(paste0("Processing sample ", sample, " ... Done!"))

}



```

## Overall stats

```{r}
ribo_file <- "/riboseq/data/processed/RiboseQC.merged/Medullo_merged/_results_RiboseQC_all"
sample <- "Medullo_merged"

# Load riboseqc data
load(ribo_file)
data <- res_all
profile_type = "profiles_P_sites"
res = "P_sites_subcodon"
comp = "nucl"
signal <- data[[profile_type]][[res]][[comp]]

# signal <- data$profiles_fivepr$five_prime_subcodon$nucl

rl_ok <- names(signal)  # get available read lengths
rl_ok <- c(rl_ok[1], as.character(sort(as.integer(rl_ok[-1], decreasing=FALSE))))  # sort read lengths in ascending order and "all" at the beginning

data_bar = list()

for (rl in rl_ok) {
  # get read length data and format
  x <- data.frame(signal[[rl]], check.names = F)
  x.m <- data.frame(value = colSums(x))
  x.m$pos <- seq_len(length(x))
  x.m$frame <- paste0("frame ", rep(c(3,1,2),67)[seq_along(x)])
  
  # add read length data to list
  data_bar[[rl]] <- x.m
}

# Plot periodicity for read length 29 nt

# Get read length for max frame preference in sample
  summary_P_sites_sample <- data.frame(data$summary_P_sites)
  rl <- summary_P_sites_sample %>%
  dplyr::slice(which.max(frame_preference)) %>%
  dplyr::select(read_length)
  rl <- as.character(rl)

# rl <- ifelse("29" %in% rl_ok, "29", "all")
# rl <- "all"
x.m <- data_bar[[rl]]

res_breaks <- c(1, 26, 51, 84, 117, 150, 176, 200)
res_labels <- c("TSS", "", "start\ncodon", "", "", "stop\ncodon", "", "TES")
res_x_lab <- "\nposition (nucleotide resolution)"

x.m$frame<-as.character(x.m$frame)
x.m$frame[x.m$frame=="frame 2"]<-"Frame 0"
x.m$frame[x.m$frame=="frame 1"]<-"Frame 2"
x.m$frame[x.m$frame=="frame 3"]<-"Frame 1"
x.m$frame<-gsub(x.m$frame,pattern = "F",replacement = "f")
x.m$frame<-factor(x.m$frame,levels=c("frame 0","frame 1","frame 2"))

p1 <- ggplot(data=x.m, aes(x=pos, xend=pos, y=0, yend=value)) +
  geom_segment(aes(colour=frame, group=frame)) +
  scale_color_manual(values = c("red","forestgreen","blue")) +
  scale_x_continuous(breaks=res_breaks, labels=res_labels) +
  #scale_y_continuous(expand=c(0, 0)) +
  labs(x=res_x_lab, y="Ribosome P-sites") +
  theme_classic() +
  theme(legend.text = element_text(size = 12),
        legend.title = element_blank(),
        axis.text = element_text(size = 10),
        axis.title = element_text(size = 12), 
        legend.position = "none")

framepref_allreads <- data_bar[[rl]] %>%
  group_by(frame) %>%
  summarize(tot_reads = sum(value)) %>%
  mutate(perc = tot_reads/sum(tot_reads))

framepref_allreads$frame<-as.character(framepref_allreads$frame)
framepref_allreads$frame[framepref_allreads$frame=="frame 2"]<-"Frame 0"
framepref_allreads$frame[framepref_allreads$frame=="frame 1"]<-"Frame 2"
framepref_allreads$frame[framepref_allreads$frame=="frame 3"]<-"Frame 1"
framepref_allreads$frame<-gsub(framepref_allreads$frame,pattern = "F",replacement = "f")
framepref_allreads$frame<-factor(framepref_allreads$frame,levels=c("frame 0","frame 1","frame 2"))

framepref_allreads_summary <- framepref_allreads_allsamples %>%
  group_by(frame) %>%
  summarize(perc_mean = mean(perc))

p2 <- ggplot(data = framepref_allreads_summary, aes(x = frame, y = perc_mean, fill = frame)) +
  geom_bar(stat = "identity") +
  ggbeeswarm::geom_quasirandom(data = framepref_allreads_allsamples, aes(x = frame, y = perc, fill = frame), cex = 0.3) +
  scale_fill_manual(values = c("red","forestgreen","blue")) +
  scale_x_discrete(labels = c("0", "+1", "+2"))+ 
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0, 1)) +
  labs(x = "Reading frame", y = "Ribosome P-sites") +
  theme_classic() +
  theme(legend.position = "none")

# p2 <- p2 +
  # ggbeeswarm::geom_quasirandom(data = framepref_allreads_allsamples, aes(x = frame, y = perc, fill = frame))


p3 <- cowplot::plot_grid(p1, p2, align = "h", axis = "bt", rel_widths = c(3, 1))

p3

cowplot::save_plot(filename = paste0("/riboseq/results/plots/", sample, "_periodicity_new.pdf"), p3, ncol = 2, base_height = 3, base_width = 4)

```


